---
title: "Length of hospital stay: Part II"
description: |
  Part 2: time-to-event analysis
author: "Chi Zhang"
date: "2024-10-02"
categories: [Survival]
sidebar: false
code-block-bg: true
code-block-border-left: true
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: false
---

> This analysis is in preparation for interviews related to time-to-event analysis. Focus will be put on the procedure (and how to do it in R), as well as interpretation of the results.

The data `liggetid` was collected at the Geriatric Department at Ullevål Sykehus. Detailed description can be found in [part 1](case_study_liggetid_1.qmd) of the case study. 

We will focus on the following variables: 

- `admission_year`, year of hospital admission
- `age`
- `sex`
- `admission_from`, where 1 = home, 2 = Div. of Medicine, 3 =
Div. of Surgery, 4 = Other division, 5 = Other hospital, 6 =
Nursing home
- `stroke`, where 1 = yes, 0 = no
- `los`, hospital stay, in days 


```{r}
#| label: data
#| echo: true
#| eval: true

los <- readRDS('data/los.rds')
head(los, 3)
```



### Simulate times and status

In this analysis we will do the following steps for processing. First we limit the analysis to data from 1985 onwards (not including).

**Simulate censor time**. We know the time for discharge is given as `los` - this is the true event time. In this example for hospital LOS, all patients are discharged (with final status 1). However *for the purpose of learning*, we randomly select 20% of the subjects and assume they are censored. The time for censored is fixed to 0.9 times their final LOS.

**Simulate stroke time**. `stroke` is a variable that is typically not measured at baseline, hence it should be treated as a time dependent variable. The time for stroke is fixed at 0.7 of their time (after censoring). 



```{r}
#| label: data-sim
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "Show the code"

suppressMessages(library(dplyr))
library(ggplot2)
suppressMessages(library(data.table))


los <- readRDS('data/los.rds')
# table(dlos$admission_from)

# to increase readability, process
data.table::setDT(los)

# code a status: 1 indicates released. all of them
los$discharge <- 1

# take those from 1985 onwards
los <- los[admission_year > 1985]

# we know the true time (los)
# now randomly select subjects (20%), make a new time, then mark censor

set.seed(1)
n <- nrow(los)
id_time <- sample(1:n, size = 0.2*n, replace = F)

los[, time := los]
los[id_time, time := round(0.9*los, digits = 0)]

# make new status
los[, status := 1]
los[id_time, status := 0] # censored

# for those with stroke, simulate a time
# if no stroke, the time is same as overall time
los[stroke == 1, time_stroke := round(0.7*time, digits = 0)]
los[stroke == 0, time_stroke := time]

head(los, 3)
```

## Kaplan-Meier analysis

Create KM curve


```{r}
#| label: km
#| echo: true
#| eval: true

library(survival)
library(ggsurvfit)

# from ggsurvfit pkg. using surfit also works
ss <- survfit2(Surv(time, status) ~1, data = los)

ggsurvfit(ss) + 
  labs(x = 'Days', 
       y = 'Overall survival prob') + 
  add_confidence_interval() + 
  add_censor_mark() + 
  add_risktable()
```


### x-day survival probability

```{r}
#| label: xdaysurv
#| echo: true
#| eval: true

# beyond 100 days
summary(survfit(Surv(time, status)~1, data = los), times = 100)
nrow(los[time>=100]) # n at risk (still alive)
# nrow(los[time<=100])  # this includes censored, 363+1
nrow(los[status == 1 & time <= 100]) # n event (already happened)
# n at risk: 38; n event 290, p = 0.15
```

401 total, 38 at risk, within the 364 (-1), 290 died 

Naive but incorrect way of calculating: 1-290/401 = 27.7%. This over-estimates how many are alive, as (364-290) = 74 are censored and we do not know the outcome.


### Median survival time

Median survival time means half of the subjects have events happen before, and half after. It corresponds to the time where survival probability is 0.5.

The median survival time is **higher** than using `median(time_to_event)` on those with outcome. It is intuitive because with censored data there are time periods that are not observed, which are ignored by `median()`. 

```{r}
#| label: median
#| echo: true
#| eval: true

survfit(Surv(time, status) ~ 1, data = los)
median(los[status == 1, time]) # only compute time for uncensored
```

Compare with the complete case when there's no censoring. We use `discharge` as the outcome status, all subjects have 1. The time variable is `los` which is the true time. Now these two should be consistent.

```{r}
#| label: median2
#| echo: true
#| eval: true

survfit(Surv(los, discharge) ~ 1, data = los)
median(los$los)
```

```{r}
#| label: km2
#| echo: true
#| eval: true

ggsurvfit(ss) + geom_vline(xintercept = c(100, 35), col = 'red')
```


::: {.callout-note collapse='true'}
## Interpretation



:::

::: {.callout-note collapse='true'}
## Risk of ignoring censored subjects

* over-estimated (higher) survival probability calculated as (1 - dead/n), since the numerator does not include censored subjects whose outcome is unknown.
* under-estimated (lower) median survival time, since their time before event will prolong the survival time.

:::





## Cox proportional hazards model




## Time varying covariate










