---
title: "Logistic regression example: lung"
description: |
  Logistic regression
author: "Chi Zhang"
date: "2024-10-05"
categories: [Regression]
sidebar: false
code-block-bg: true
code-block-border-left: true
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: false
---

> This analysis is in preparation for interviews related to logistic regression. Focus will be put on the procedure (and how to do it in R), as well as interpretation of the results.




```{r}
mtcars |> head()

mlr1 <- glm(vs ~ mpg + wt, family = 'binomial', data = mtcars)
summary(mlr1)
```



Deviance

* null deviance: 43.8, when only intercept is fitted to the data
* residual deviance: 25.3, when model has two predictors

The difference is how much adding two variables has improved the model

## Hypothesis tests

Wald test: z-values are the wald test statistics

### Likelihood ratio test for two nested LR

Build a second model 


```{r}

mlr2 <- glm(vs ~ mpg, family = 'binomial', data = mtcars)
summary(mlr2)
anova(mlr2, mlr1, test = 'Chisq') # analysis of deviance
```



```{r}
mlr3 <- glm(vs ~ wt, family = 'binomial', data = mtcars)
summary(mlr3)
anova(mlr3, mlr1, test = 'Chisq') # analysis of deviance
```



```{r}
AIC(mlr1, mlr2, mlr3)
BIC(mlr1, mlr2, mlr3)
```

## Diagnostics








