---
title: "Interview: statistics used in my work"
description: |
  Case studies
author: "Chi Zhang"
date: "2024-10-20"
categories: [Inference]
sidebar: false
code-block-bg: true
code-block-border-left: true
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: false
---

> Interview preparation: statistics and study designs used in my own work. Divide by types of endpoint, design (and why), analysis method.


<details>
  <summary>Selected list of my statistical advising work</summary>
  
### Observational studies
  
| Study | Endpoint| Design  | Analysis | Comment |
|-------|-------|------|------|------|
| Hansen 2018 |  QoL on vocal surgical questionnaire, likert scale  | Case control, 3 groups: controls (no voice problem), test-retest group, surgical group   |  Cohen's kappa   |   Reliability of questionnaire   |
| Sarjomaa 2022 |    |    |     |  Tab 4    |
| Sarjomaa 2024 |    | Case control (with population)   |     |      |
| Kabashi 2019 |    | Cross-sectional, multi-site   |     |   Tab 2 and 3   |
| Fell 2024 |  CB prevalence  |    |     |      |
| Diab 2023 | ECG parameter. Pressure measures   |  n=20  | t-test or wilcoxon, correlation    |  Choice of method    |

### Clinical trials

| Study |Endpoint & Tx | Design  | Analysis | Comment |
|---------|---| -----|------|------|
| [Reier-Nilsen 2019](https://doi.org/10.1111/pai.13066) | QoL, Oral immunotherapy |  RCT, 2 year follow-up (T0, T1, T2). 2:1 block.   |     |      |
| Opdal 2023 | QT intervals. Methadone vs R-methadone   | n=9. SS based on one-sample t-test with effect size (QT in ms) 20/15   |     |   SS calculation   |
| Simonsen 2023 | QoL, stigma stress scale   | n=32, 1:1. Three endpoints (T0, T1, T2)   |     |  Revise study protocol, SS    |
| Callender 2023 |  Artery measures for blood flow  | n=26. Two legs, one tx one ct. Tx has 3 levels of stimulus. Within-subject design   |  Anova; or mixed model   |  Choice of method    |

Pay attention to how the sample size is calculated


</details>


For behavior questions check [here](interview_behavior.qmd)


::: {.callout-note collapse='true'}
## The kind of studies you've done

* by field: Covid; nutrition; 
* by design: 
* by endpoint: most are *QoL*

:::


::: {.callout-note collapse='true'}
## What do you like and dislike about the work

Like:  

* like to help clinicians understand their data
* like health data, sometimes you see very novel data types (e.g. measurements by the british doctor)
* interesting ways people carry out their studies

Dislike:

* communication (e.g. p-value interpretation)
* sometimes feels the frustration that the importance of data quality, processing is undervalued, and we are just p-value producers
* often need to re-work multiple rounds

:::

::: {.callout-note collapse='true'}
## Challenges you had in a study


:::


# Endpoints

### Quality of life QoL

Typically measured on a scale, summed up 

* QoL on vocal surgical questionnaire, likert scale. Intervention: throat surgery (Hansen 2018)
* QoL on likert scale for both children and parents. Intervention: oral immunotherapy (Reier-Nilsen 2019)
* QoL, stigma stress scale (Simonsen 2023)


### Biomarker

* Artery blood flow measures. Intervention: local anaesthesia medicine (Callender 2023)
* QT intervals. Intervention: methadone (Opdal 2023)
* ECG parameters. Intervention: pressure measures (Diab 2023)


### Observational: Covid / CB


# Design

## RCT

> Parallel or cross-over? Choice of control - placebo, active control, historical control? Timing and frequency of measurements? Randomization method and blinding?


My works have an exploratory nature - the RCT works I've done all seems to be phase II (neither determining the safe dose, nor find deterministic evidence in the wider population) with small to medium sample size.

I have never followed an RCT from start to end, so I've never been involved in the design of the studies.

For the designs: mostly parallel designs with **single measurement** or **longitudinal measurements**, where subjects are measured multiple times. I've done one **cross-over** study (same patient, first tx then ct).

I have not done any factorial design (multiple treatments at the same time).

Selection criteria: clinicians have their own criteria for relevant subjects, usually age groups, sex and underlying conditions

Randomization: the ones I've done are mostly **simple randomization**, nothing with blocks or stratification

Masking (blinding): my works seem to have open labels, not blinded 


#### Peanut allergy

(Reier-Nilsen 2019)

2 year follow up (T0, T1, T2)

* Endpoint: 3 or 5-point likert scale for kids, 7 point likert scale for parents. Points are summed and reported
* Selection: Enroll positive food challenge 5-15 yo, 96 in total. Exclude asthma and other severe chronic diseases
* Randomization: Remove 19 ineligible, 77 2:1 block OIT vs controls (observation only). Possible reason for 2:1: many would drop out
* Blinding: NA


#### Methadone

(Opdal 2023) 

* Selection: 18+, not pregnant, no serious illness, willingness to comply
* Design: Cross over, 14 days washoff
* Randomization: Cross over, subjects are their own controls
* Blinding: Cardiologist is blinded to calculate QT measurement
* Comparison: on-methadone QTc vs off
* Sample size: n=9, clinically relevant difference 20, SD 15 (unit: microsecond), b = 0.9. Can be computed with `pwr::pwr.t.test(d=20/15)`. See more in the next section.


#### Artery blood flow

(Callender 2023) n = 26; within-subject two legs; three levels of intervention (stimulus)

* Selection: 13M, 13F, healthy, between age 18-50, free from medication influencing cardiovascular system
* Design: Parallel
* Randomization: two legs on the same patient, tx ct
* Blinding: NA
* Sample size: a=0.05, b=0.8; effect size based on previous works, computed by clinicians themselves




## Power and sample size

> clinicians provide information on the anticipated effect size, statistician calculate sample size needed to detect effect with adequate power while minimizing type I and II error.


Sample size computation: 

* Require:  
  * test: t-test (one, two samples), tails
  * significance level
  * power
  * cohen's d or f - effect size (e.g. standardized mean difference). If unavaiable, can use the Cohen's guideline for small, medium and large effect sizes: (f = 0.1, 0.25, 0.4)


#### QoL stigma

(Simonsen 2023) n = 32; T0, T1, T2; 1:1; justify SS computation

If the study has a paired design with effect size (standardized after substracting the standard deviation) 0.5,

```{r}
#| label: ss1
#| warning: false
#| echo: true

pwr::pwr.t.test(d = 0.5, sig.level = 0.05, power = 0.8, type = 'one.sample', alternative = 'two.sided')
```

#### Methadone

(Opdal 2023) n=9; effect size provided (as clinically relevant difference and standard deviation); compute SS

In this case study, the computed sample size is as follow, 

```{r}
#| label: ss2
#| warning: false
#| echo: true

pwr::pwr.t.test(d = 20/15, sig.level = 0.05, power = 0.9, type = 'one.sample', alternative = 'two.sided')
```

The effect size is huge, hence the small sample size required. 

#### Post-hoc power computation 

It is not recommended, as the information is the same as what p-value provides: a small p-value that indicates a large effect size, that will give a high power.

Power analysis is a design tool, not a diagnostic tool. It is used to make sure that the study has enough samples to detect an effect of a given magnitude and the study has high probability of detecting a true effect if it exists. Once the study is done, analyzing power adds little beyond the p-value.




## Observational

> Eligibility criteria?

#### AHUS 

Selected as subjects have been in the hospital during a certain time, above a certain age and have been in certain wards

#### COVITA

(Sarjomaa 2024) Case control with population 


#### Alcohol

(Kabashi 2019) Cross sectional, multi-site

## Survey

#### Norkost





# Analysis




## Tests

> Superiority, equivalency and non-inferiority


### t-test, wilcoxon

### Multiple testing adjustment

#### NAV paper


## Regression

### Linear regression / regularization

#### HAI paper

#### FHI Excess mortality



### Logistic regression / ordinal

#### DTW paper

LR as a classifier

#### COVITA




### Poisson regression

#### Transfer paper

Transfer rate 

#### NAV

Over-dispersion


## Missing data

> Complete case only? Imputation? LOCF (last observation carried forward)? IIT vs per protocol

#### (not typical) DTW



#### (not typical) Norkost dietary survey

Keep zero as is


## Subgroup analysis, sensitivity 

#### Transfer paper


#### CB


## Mixed model

#### Peanut allergy

#### Norkost

## Survival analysis

#### Hospital LOS


## Others 

Cohen's kappa




