---
title: "Sample size (part II)"
description: |
  Regression
author: "Chi Zhang"
date: "2023-08-20"
categories: [Sample size]
sidebar: false
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: false
---


(This is the part II on sample size calculation)





# Correlation

Correlation coefficient is used for effect size measure. Usse 0.1, 0.3, 0.5 to represent small, medium and large sizes.

```{r}
#| label: corr
#| warning: false
#| echo: true

# correlation coeff
pwr::pwr.r.test(r = 0.3, sig.level = 0.05, power = 0.8, alternative = 'two.sided')
```



# Linear regression (F-test)

F-test for linear regression is testinng whether $R^2$ is greater than zero (one-sided). $R^2$ is the explained variance by using the predictors, $R^2 = 0.3$ means that 30% of the variance are explained by the model.

Cohen's f2, based on $R^2$, goodness of fit ($f2 = R^2/(1-R^2)$). use 0.02, 0.15, 0.35 to represent small, medium and large effect sizes.

* u: number of predictors
* v: n-u-1
* as a result, sample size n = v+u+1

```{r}
#| label: lm
#| warning: false
#| echo: true

# effect size f2 = 0.15; use u=3 predictors
pwr::pwr.f2.test(u = 3, f2 = 0.3, sig.level = 0.05, power = 0.8)
```

Here v = 73, sample size is 73+3+1 = 77.

Alternatively, can use `pwrss::pwrss.f.reg()`. The parameter is r2 rather than f2 (but can also use f2).

```{r}
#| label: lm-2
#| warning: false
#| echo: true

pwrss::pwrss.f.reg(r2 = 0.3, k = 0.3, power = 0.8, alpha = 0.05)
# should be equivalent to
# pwr::pwr.f2.test(u = 3, f2 = 0.3/0.7, sig.level = 0.05, power = 0.8)
```




# GLM









### Resources 

* Sample size calculation in clinical trial using R.  
Park et al. 2023. https://doi.org/10.7602/jmis.2023.26.1.9

* Bulus, M (2023) pwrss: Statistical Power and Sample Size Calculation Tools. R package version 0.3.1. https://CRAN.R-project.org/package=pwrss. [Vignette documentation](https://cran.r-project.org/web/packages/pwrss/vignettes/examples.html)




