---
title: "Sample size (part I)"
description: |
  Overview, mean and proportion comparison
author: "Chi Zhang"
date: "2023-08-20"
categories: [Sample size]
sidebar: false
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    code-tools: false
---

(This is the part I on sample size calculation)


Sample size calculation is to determine the **smallest** number of subjects required, to detect a clinical meaningful effect. Why not recruiting as many as possible? Too expensive; or unethical (i.e. more people will be having potentially harmful or futile treatments). 

## Relevant concepts

Study design

* parallel: group 1 TxA, group 2 TxB
* crossover: requires fewer ssample than parallel; but requires wash-out period. Group 1 TxA -> TxB; group 2 TxB -> TxA 

Tests

* $\mu_T, \mu_s$: mean of new Tx or standard procedure
* $\delta$: minimum clinically important difference
* $\delta_{NI}$: non-inferiority margin

| Test for |  H0    |   H1    | 
|:--------:|:------:|:-------:|
| Equality | $\mu_T - \mu_s = 0$| $\mu_T - \mu_s \neq 0$ |
| Equivalence | $|\mu_T - \mu_s| \geq 0$ | $|\mu_T - \mu_s| < 0$ | 
| Superiority | $\mu_T - \mu_s \geq 0$ | $\mu_T - \mu_s < 0$ | 
| Non-inferiority | $\mu_T - \mu_s \leq -\delta_{NI}$ |  $\mu_T - \mu_s \geq -\delta_{NI}$| 


Errors

* Type I error, significance level $\alpha$. P(reject H0 |H0). Usually set to 0.05
* Type II error $\beta$. P(not reject H0 |H1). 
* Power, $1 - \beta$. P(reject H0 |H1). Usually set to 80% or 90%

Primary outcome

* can be categorical or continuous
* Minimal meaningful detecable difference MD: the smallest difference to be considered as clinically meaningful in the primary outcome

Dropout rate: need to be adjusted. 

Allocation ratio: unequal sample size. 

Effect size (Cohen's d, f) should be found in the literature. In general, 

* very small, d = 0.01
* small, d = 0.2
* medium, d = 0.5
* large, d = 0.8
* very large, d = 1.2
* huge, d = 2


# Continuous outcome

Parametric: t-tests, ANOVA

Non-parametric: Wilcoxon tests, Kruskal-Wallis

A rule of thumb for the sample size of non-parametric tests: compute the parametric alternative, then add 15%.

## One sample t-test

```{r}
#| label: ttest-onesample
#| warning: false
#| echo: true

# effect size: 0.5
pwr::pwr.t.test(d = 0.5, sig.level = 0.05, power = 0.8, type = 'one.sample', alternative = 'two.sided')

# command for paired t-test is the same; but d is computued differently
# pwr::pwr.t.test(d = 0.5, sig.level = 0.05, power = 0.8, type = 'paired', alternative = 'two.sided')
```


## Two sample t-test

Note that this result is for one group: in total it's times two.

```{r}
#| label: ttest-twosample
#| warning: false
#| echo: true

# effect size: 0.5
pwr::pwr.t.test(d = 0.5, sig.level = 0.05, power = 0.8, type = 'two.sample', alternative = 'two.sided')
```


## ANOVA

Result is for each group.

```{r}
#| label: anova
#| warning: false
#| echo: true

# k: number of groups; f: effect ssize
pwr::pwr.anova.test(k = 3, f = 0.25, sig.level = 0.05, power = 0.8)
```



# Categorical outcome

## Proportions

Cohen's h is used as the effect size, $h = 2arcsin(\sqrt{p_1} - 2arcsin(\sqrt{p_2}))$. Use 0.2, 0.5, 0.8 for small, medium and large effect sizes.


```{r}
#| label: proportions
#| warning: false
#| echo: true

# one group
pwr::pwr.p.test(h = 0.5, sig.level = 0.05, power = 0.8, alternative = 'two.sided')

# two groups
pwr::pwr.2p.test(h = 0.5, sig.level = 0.05, power = 0.8, alternative = 'two.sided')
```


## Chi-square test

Cohen's w. Use $l, k$ to compute degrees of freedom.

```{r}
#| label: chisq
#| warning: false
#| echo: true

# k: number of groups; f: effect ssize
pwr::pwr.chisq.test(w = 0.3, df = (2-1)*(3-1), sig.level = 0.05, power = 0.8)
```


## Exact test

Need to specify the proportion in each group (control, treatment). Allocation ratio 1:1

```{r}
#| label: exact2x2
#| warning: false
#| echo: true

exact2x2::ss2x2(p0 = 0.2, p1 = 0.8, n1.over.n0 = 1, sig.level = 0.05, power = 0.8, 
                approx = F, print.steps = T, paired = F)
```






# Correlation


# Linear regression 


# GLM









### Resources 

* Sample size calculation in clinical trial using R.  
Park et al. 2023. https://doi.org/10.7602/jmis.2023.26.1.9

* Bulus, M (2023) pwrss: Statistical Power and Sample Size Calculation Tools. R package version 0.3.1. https://CRAN.R-project.org/package=pwrss. [Vignette documentation](https://cran.r-project.org/web/packages/pwrss/vignettes/examples.html)
